---
title: "Home Schooling in the United States"
output: 
  flexdashboard::flex_dashboard:
    theme: yeti
    social: menu
    orientation: columns
    vertical_layout: fill
engine: shiny
runtime: shiny
---

```{r setup, include=FALSE}
library(flexdashboard)
library(lubridate)
library(dplyr)
library(ggplot2)
library(ggstance)
library(ggalt)
library(plotly)
library(shiny)
library(highcharter)
#library(scales)

# Read the dataset
schools <- read.csv('https://raw.githubusercontent.com/washingtonpost/data_home_schooling/main/home_school_district.csv')

# Remove the second column (you mentioned -2)
schools <- schools[, -2]

# Function to extract the first year from the academic year
extract_first_year <- function(academic_year) {
  first_year <- as.numeric(strsplit(academic_year, "-")[[1]][1])
  return(first_year)
}

# Apply the extraction function to the 'year' column
schools$year <- sapply(schools$year, extract_first_year)
schools$homeschool_students[is.na(schools$homeschool_students)] <- 0

str(schools)
# Group by year and calculate the sum of homeschool_students
schools_summarized <- schools %>%
  group_by(year) %>%
  summarise(total_homeschool_students = sum(homeschool_students, na.rm = TRUE))


str(schools_summarized)

# Assuming your data frame is named 'schools'
top_states <- schools %>%
  group_by(state) %>%
  summarise(total_homeschool_students = sum(homeschool_students, na.rm = TRUE)) %>%
  arrange(desc(total_homeschool_students)) %>%
  head(5)


all_states <- unique(schools$state)
all_districts <- unique(schools$lea_name)

```

Inputs {.sidebar}
==================================

The Washington Post is compiling a database of from individual school districts and states as a whole regarding home-school enrollment from 2017-18 through 2022-23 in the United States and has made that database [publicly available on GitHub](https://github.com/washingtonpost/data_home_schooling/). Their own visualizations and reporting are [online here](https://www.washingtonpost.com/education/interactive/2023/homeschooling-growth-data-by-district/), and this dashboard presents the same data. 

Author : [Soham Bhagwat](sohambhagwat.tech)

----------------------------------

```{r}
selectInput("state_dropdown", label = "Select State", choices = all_states, selected = all_states[1])
selectInput("district_dropdown", label = "Select School District", choices = all_districts, selected = all_districts[1])

observeEvent(input$state_dropdown, {
  selected_state <- input$state_dropdown
  filtered_districts <- schools %>% filter(state == selected_state)
  
  # Update the district dropdown
  updateSelectInput(session, "district_dropdown", choices = unique(filtered_districts$lea_name))
})


```

Insights {data-icon="fa-users"} 
=====================================  

Row
-------------------------------------
   
### Home schooled children over the years

```{r}
p <- ggplot(schools_summarized, aes(x = year, y = total_homeschool_students)) +
  geom_line(color = "blue") +
  geom_point(color = "red") +
  labs(x = "Year",
       y = "Total Homeschooled Children") +
  scale_y_continuous(labels = scales::label_number(scale = 1e-3, suffix = "k")) +
  theme_minimal()

# Convert ggplot to interactive plot using plotly
p <- ggplotly(p)

p

```

Column {data-width=350}
-----------------------------------------------------------------------

### Top 5 states with highest homeschooling

```{r}
output$bar_chart_states <- renderHighchart({
  highchart() %>%
    hc_chart(type = "bar") %>%
    hc_xAxis(categories = top_states$state) %>%
    hc_yAxis(title = list(text = "Total Homeschooled Children")) %>%
    hc_tooltip(valueDecimals = 0) %>%
    hc_plotOptions(bar = list(dataLabels = list(enabled = TRUE))) %>%
    hc_colors("#3498db") %>%
    hc_add_series(name = "Total Homeschooled Children", data = top_states$total_homeschool_students)
})
highchartOutput("bar_chart_states")
```

### Homeschooled children by School Districts for a given State.

```{r}
# Define a reactive expression to filter data based on user inputs
filtered_data <- reactive({
  selected_state <- input$state_dropdown
  selected_district <- input$district_dropdown
  schools %>%
    filter(state == selected_state, lea_name == selected_district)
})

library(highcharter)
# Create an interactive bar chart for the selected school district
output$bar_chart_district <- renderHighchart({
  highchart() %>%
    hc_chart(type = "column") %>%
    hc_title(text = paste("State:", input$state_dropdown, ", School District:", input$district_dropdown)) %>%
    hc_xAxis(categories = filtered_data()$year) %>%
    hc_yAxis(title = list(text = "Homeschooled Children")) %>%
    hc_tooltip(valueDecimals = 0) %>%
    hc_add_series(name = "Homeschooled Children", data = filtered_data()$homeschool_students, colorByPoint = TRUE) %>%
    hc_plotOptions(column = list(dataLabels = list(enabled = TRUE, color = "white"))) %>%
    hc_colors("#3498db") %>%
    hc_legend(enabled = FALSE)
})

highchartOutput("bar_chart_district")
```

About {data-icon="fa-question"} 
=====================================  
This application was developed by [Soham Bhagwat](sohambhagwat.tech) to find the trends in the homeschooling in the United States.

The data is collected by [Washington Post](https://www.washingtonpost.com/) and is available [here](https://github.com/washingtonpost/data_home_schooling/).

The application is built with the [Shiny](http://shiny.rstudio.com) framework for the [R programming language](https://www.r-project.org/). The application layout is produced with the [flexdashboard](http://rstudio.github.io/flexdashboard/index.html) package, and the charts and maps use [Plotly](http://plot.ly), [Highcharts](http://www.highcharts.com/), and [ggplot2](http://ggplot2.org/), all accessed through their corresponding R packages.


To learn more about my work, [visit my website](https://sohambhagwat.tech) or [connect with me on Linkedin](https://www.linkedin.com/in/soham-bhagwat/).

I'm actively looking for entry level Data Scientist and Data Engineer roles and my resume can be found [here](https://www.sohambhagwat.tech/pdf/Soham_Bhagwat.pdf) or the best way to reach me is via [email](sohambhagwat2@gmail.com) 


